### 2010-07-26 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix the Env widow line
Desc@ICFP: re-shaping inconsistency
Desc@ICFP: Jack the Ripper at work. Another victim.

In the previous section, we have presented a language of closed
arithmetic expressions. Using the free monad construction, we are
going to extend this construction to open terms. An open term is
defined with respect to a context, represented by a snoc-list of
types:
%
\[
\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
\Ctxt           & \Set \\
\SYMBCtxtEmpty  & \Ctxt \\
\SYMBCtxtSnoc   & \Ctxt \To \Ty \To \Ctxt
\end{array}
\]
%
An environment realises the context, associating a value for each type:
%
\[
\stk{
\SYMBEnv : \Ctxt \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Env{\CtxtEmpty}                & \Unit \\
\Env{(\CtxtSnoc{\V{G}}{\V{S}})} & \TIMES{\Env{\V{G}}}{\Val{\V{S}}}
\end{array}
}\]
%
In this setting, we define $\SYMBVar$ by:
%
\[\stk{
\Var{}{} : \Ctxt \To \Ty \To \Set \\
\begin{array}{@{}ll@{\:\:\mapsto\:\:}l}
\Var{\CtxtEmpty}{& \V{T}}                & 
    \Void \\
\Var{(\CtxtSnoc{\V{G}}{\V{S}})}{& \V{T}} & 
    \SUM{(\Var{\V{G}}{\V{T}})}{(\V{S} \PropEq \V{T})}
\end{array}
}\]
%
While $\SYMBVal$ maps the type to the corresponding host type,
$\SYMBVar$ indexes a value in the context, obtaining a proof that the
types match. The $\SYMBlookup$ function precisely follow this
semantics:
%
\[\stk{
\SYMBlookup : \PI{\V{G}}{\Ctxt} 
          \Env{\V{G}} \To 
          \PI{\V{T}}{\Ty} 
          \Var{\V{G}}{\V{T}} \To
          \Val{\V{T}} \\
\begin{array}{@{}l@{}l@{}l@{}l@{}lll}
\lookup{& (\CtxtSnoc{\V{G}}{.T})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumRight{\C{refl}})} & \mapsto & \V{t} \\
\lookup{& (\CtxtSnoc{\V{G}}{\V{S}})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumLeft{\V{x}})} & \mapsto & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
\end{array}
}\]

@@

Consequently, taking the free monad of \(\HExprD\) by \(\SYMBVar\:
\V{G}\), we obtain the language of open terms in a context \(\V{G}\):
%
\[
\openTerm{\V{G}} \mapsto \FreeIMonad{\HExprD}{\Ty}{(\SYMBVar\:\V{G})}
\]
%
In this setting, the language of closed terms corresponds to the free
monad assigning an empty set of values to variables
%
\[
\closeTerm \mapsto \FreeIMonad{\HExprD}{\Ty}{\SYMBEmpty}
\quad
\mbox{where}
\quad
\stk{
\SYMBEmpty : \Ty \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Empty{\_}   & \Zero \\
\end{array}
}\]
%
Allowing variables from an empty set is much like forbidding variables,
so \(\closeTerm\) and \(\HExprD\) describe isomorphic
datatypes. Correspondingly, you can update your old \(\HExprD\) to a
shiny \(\closeTerm\) with \(\update\):
%
\[\stk{
\update : \muide{\Ty}{\HExprD} \DotTo \muide{\Ty}{\closeTerm} \\
\begin{array}{@{}l@{}l}
\update\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty}\: & (\toIDesc{\HExprD})\:
                                                  (\muide{\Ty}{\closeTerm})\: \\
                                                & (\LAM{\_} \LAM{\pair{\V{tag}}{\V{tm}}{}} \Con{\pair{\Su{\V{tag}}}{\V{tm}}{}})\:
                                                  \V{ty}\:
                                                  \V{tm}
\end{array}
}\]
% 
The other direction of the isomorphism is straightforward, the
$\SYMBDVar$ case being impossible. Therefore, we are entitled to
reuse the $\evalH$ function to define the semantics of $\closeTerm$.

@@

The fact that indexed descriptions are closed under substitution
is potentially of considerable utility, if we can exploit this fact:
\[
\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
    \mapsto 
        \idescop{\V{D}}
                {\V{I}}
                {\LAM{\V{i}}
                     {\idescop{\V{\sigma} \V{i}}
                              {\V{J}}
                              {\V{X}}}}
        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
\]
By observing that a description can be decomposed via substitution, we
split its meaning into a superstructure of substructures, e.g. a
`database containing salaries', ready for traversal operations
preserving the former and targeting the latter. We shall now turn to
some practical use of the free monad construction.

>>>>

In the previous section, we have presented a language of closed
arithmetic expressions. Using the free monad construction, we are
going to extend this construction to open terms. An open term is
defined with respect to a context, represented by a snoc-list of
types:
%
\[
\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
\Ctxt           & \Set \\
\SYMBCtxtEmpty  & \Ctxt \\
\SYMBCtxtSnoc   & \Ctxt \To \Ty \To \Ctxt
\end{array}
\]
%
An environment realises the context, packing a value for each type:
%
\[
\stk{
\SYMBEnv : \Ctxt \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Env{\CtxtEmpty}                & \Unit \\
\Env{(\CtxtSnoc{\V{G}}{\V{S}})} & \TIMES{\Env{\V{G}}}{\Val{\V{S}}}
\end{array}
}\]
%
In this setting, we define $\SYMBVar$ by:
%
\[\stk{
\Var{}{} : \Ctxt \To \Ty \To \Set \\
\begin{array}{@{}ll@{\:\:\mapsto\:\:}l}
\Var{\CtxtEmpty}{& \V{T}}                & 
    \Void \\
\Var{(\CtxtSnoc{\V{G}}{\V{S}})}{& \V{T}} & 
    \SUM{(\Var{\V{G}}{\V{T}})}{(\V{S} \PropEq \V{T})}
\end{array}
}\]
%
While $\SYMBVal$ maps the type to the corresponding host type,
$\SYMBVar$ indexes a value in the context, obtaining a proof that the
types match. The $\SYMBlookup$ function precisely follow this
semantics:
%
\[\stk{
\SYMBlookup : \PI{\V{G}}{\Ctxt} 
          \Env{\V{G}} \To 
          \PI{\V{T}}{\Ty} 
          \Var{\V{G}}{\V{T}} \To
          \Val{\V{T}} \\
\begin{array}{@{}l@{}l@{}l@{}l@{}lll}
\lookup{& (\CtxtSnoc{\V{G}}{.T})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumRight{\C{refl}})} & \mapsto & \V{t} \\
\lookup{& (\CtxtSnoc{\V{G}}{\V{S}})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumLeft{\V{x}})} & \mapsto & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
\end{array}
}\]

@@

Consequently, taking the free monad of \(\HExprD\) by \(\SYMBVar\:
\V{G}\), we obtain the language of open terms in a context \(\V{G}\):
%
\[
\openTerm{\V{G}} \mapsto \FreeIMonad{\HExprD}{\Ty}{(\SYMBVar\:\V{G})}
\]
%
In this setting, the language of closed terms corresponds to the free
monad assigning an empty set of values to variables
%
\[
\closeTerm \mapsto \FreeIMonad{\HExprD}{\Ty}{\SYMBEmpty}
\quad
\mbox{where}
\quad
\stk{
\SYMBEmpty : \Ty \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Empty{\_}   & \Zero \\
\end{array}
}\]
%
Allowing variables from an empty set is much like forbidding variables,
so \(\closeTerm\) and \(\HExprD\) describe isomorphic
datatypes. Correspondingly, you can update an \(\HExprD\) to a
\(\closeTerm\):
%
\[\stk{
\update : \muide{\Ty}{\HExprD} \DotTo \muide{\Ty}{\closeTerm} \\
\begin{array}{@{}l@{}l}
\update\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty}\: & (\toIDesc{\HExprD})\:
                                                  (\muide{\Ty}{\closeTerm})\: \\
                                                & (\LAM{\_} \LAM{\pair{\V{tag}}{\V{tm}}{}} \Con{\pair{\Su{\V{tag}}}{\V{tm}}{}})\:
                                                  \V{ty}\:
                                                  \V{tm}
\end{array}
}\]
% 
The other direction of the isomorphism is straightforward, the
$\SYMBDVar$ case being impossible. Therefore, we are entitled to
reuse the $\evalH$ function to define the semantics of $\closeTerm$.

@@

The fact that indexed descriptions are closed under substitution
is potentially of considerable utility, if we can exploit this fact:
\[
\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
    \mapsto 
        \idescop{\V{D}}
                {\V{I}}
                {\LAM{\V{i}}
                     {\idescop{\V{\sigma} \V{i}}
                              {\V{J}}
                              {\V{X}}}}
        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
\]
By observing that a description can be decomposed via substitution, we
split its meaning into a superstructure of substructures, e.g. a
`database containing salaries', ready for traversal operations
preserving the former and targeting the latter.
